Fractional Types

### 2013-02-20 frac.tex
major changes

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that $a$ and $b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

@@

A major open problem remains, however: a higher-order extension
of {{Pi}}. This extension is of fundamental importance in all the
originating research areas. In physics, it allows the process itself or the
observer to be treated as ``data'' that can be transformed by higher-level
processes or observed by meta-level observers. In mathematics and logic, it
allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
to itself be expressed as an isomorphism (of a higher type)
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code.

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{compact closed category} from the underlying monoidal
category for {{Pi}}. Although several such constructions that work in special
cases are known in the literature (surveyed below), there are many subtleties
in adapting these ideas to {{Pi}}. Our main novel technical device to
achieving the higher-order extension is a \emph{fractional type}, which we so
named because of its duality with conventional product types.  By adding a
formal (linear) dual to (type theoretic) products, we can internalize the
notion that ``taking an input'' corresponds to negative information, which
turns out to be the basis of first-class functions. The remainder of the
paper reviews {{Pi}} and then introduces the syntax and semantics of the
extension with fractional types. We then study the properties of the extended
language and establish its expressiveness via several constructions and
examples that exploit its ability to model higher-order computations.

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent Types include the unit type {{1}}, sum types {{b1+b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary.  In particular we do not treat the above
values as first-class citizens.  They only occur when we want to observe the
effect of an isomorphism.

\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
(value) type isomorphisms.  They are also typed, by the shape of the (value)
type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
\noindent Each line of the above table introduces a pair of dual
constants\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

Mathematically, {{Pi}} can be interpreted in the category of finite sets and
bijections: each value type denotes a finite set of the corresponding size
and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{Pi}} is given
using two mutually recursive interpreters: one going forward and one going
backwards. The use of {{sym}} switches control from one evaluator to the
other. We will present the operational semantics in the next section along
with the extension with fractional types and values. For now, we state
without proof that the evaluation of well-typed combinators always terminates
and that {{Pi}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

The relational semantics can be given an operational interpretation by
defining two functions: a forward evaluator and a backwards evaluator that
exercise the relation in opposite directions. It follows that if the forward
evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer. More
generally, the semantics of {{PiTF}} using sets and relations, i.e., in the
category of finite sets and relations suggests a few additional immediate
properties.  The category of finite sets and relations is an example of a
\emph{biproduct dagger compact closed category} which are the categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value. We return to
such properties of {{PiTF}} in Sec.~\ref{sec:cat} after we present
programming examples that illustrate {{PiTF}} and its ability to express
various higher-order notions.

@@

\begin{comment}
% We are not doing groupoids in this paper
Each type in {{PiTF}} is interpreted as a \emph{groupoid}. A groupoid can be
equivalently viewed as an oriented graph, a generalization of a group, or a
special category. We present the categorical view below.

Fractional types and values add considerable expressiveness to our
language. 
\end{comment}
In the following, we use {{bool}} as an abbreviation of {{1+1}} 
with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
abbreviation for {{right ()}}.

@@

As motivated, with fractional types, we can define a value that denotes a
first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
and a value {{v2 : b2}}. In other words, it corresponds to a function or
relation which when given matched with a value {{v1 : b1}} ``releases'' the
value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:

@@

As a simple example, consider combinator {{not}} denoting boolean negation.
It is straightforward to check that the combinator 
{{name not : 1 <-> (bool :-* bool)}} denotes the following
relation:

%subcode{proof} include main
%@ ~
%@@ () ~~(name not)~~ (1/false,true)
%
%@ ~
%@@ () ~~(name not)~~ (1/true,false)

\noindent The relation {{name not}} can therefore be used to convert any
{{false}} to {{true}} and vice-versa. We formalize this ``apply'' combinator
below.

@@

We are now a small step to implementing various higher-order
combinators that manipulate functions or relations. In particular, we
can \emph{apply} and \emph{compose} values representing relations as
shown below:

@@

\jacques{in the current paper, this seems to no longer make sense}
\begin{example}[Annihilation]
\label{ch3:ex;annihilation}
Consider the circuit below:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}
As motivated earlier, it is clearly impossible for the constraint implied by
{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
its negation. Thus there is no incoming value of type {{b}} that can ever go
through the circuit. Indeed, if we call the circuit above {{c}}, then we can
calculate that {{c}} does not relate {{v}} to anything. We call such a
circuit an annihilation circuit.
\end{example}

@@

Relational programming leverages set-theoretic relations and their
composition to express computational tasks in a declarative way. For
instance, consider this example:

{{parent = {[ (A,B), (B,C), (A,D) ]} }}

{{grandparent = parent (o) parent}}

\noindent The example defines a relation {{parent}} specified using a
set of tuples and another relation {{grandparent}} specified using the
composition of two parent relations. If we wrote this example in a
relational language (e.g., Prolog) and we executed the query
{{grandparent(A)}}, we would get the answer {{ {[C]} }}.

It turns out that with the addition of the multiplicative trace, and
the move to relations motivated in the previous section, we can
express relational programming. We illustrate the idea using a
complete example.

@@

The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{trace* cInner}} is the desired combinator
{{c_R}}. A little experimentation shows that we get the desired behavior
if {{cInner}} is chosen to behave as follows:

%subcode{proof} include main
%@ ~
%@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
%----
%@ ~
%@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
%----
%@ ~
%@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
%----
%@ ~
%@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
%----
%@ ~
%@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
%----
%@ ~
%@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
%----
%@ ~
%@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
%----
%@ ~
%@@ ((true,true),true) ~~cInner~~ ((false,false),true)

In the first three lines, the first argument is a fixed point and hence the
multiplicative trace would map the second input to the second output
producing the desired relation. In the remaining five cases, the first
argument is not a fixed point and hence all these cases would be rejected as
solutions to the constraint imposed by the multiplicative trace. It simply
remains to find the actual realization of {{cInner}} that would produce the
behavior above. One can infer the following implementation:

@@

The above example should convince the reader that the language {{Pi}}
with multiplicative trace is expressive enough to model finite relational
programming. We will not prove this claim. We develop an even more
substantial example: a SAT solver.

%%%%%%%%%%%%%%%
\subsection{Solving Constraints}
\label{ch3:sec:constraints}

A large class of constraint satisfaction problems can be expressed
using multiplicative traces. We illustrate the main ideas with the
implementation of a SAT solver.

In the usual setting, an instance of SAT is a function~{{f}} which,
when given some boolean inputs, returns {{true}} or {{false}}. The
function returns {{true}} when the inputs satisfy the constraints
imposed by the structure of {{f}} and a solution to the SAT problem is
the set of all inputs on which {{f}} produces {{true}}. The basic idea
of our construction is to generalize the annihilation circuit from
Ex.~\ref{ch3:ex;annihilation} to only annihilate values that fail to
satisfy the constraints represented by the SAT instance {{f}}. To
achieve this goal, we must however deal with several important
details.

First, because we are in a reversible world, our instance of SAT must be
expressed as an isomorphism: this is easily achieved by the construction
which embeds any boolean function {{f}} into a reversible one {{iso_f}} with
a larger domain and range.  Given such a reversible function {{iso_f}} which
represents a SAT instance, we first construct the circuit below:

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/sat2.pdf}
}
\end{center}

As shown in the circuit, the reversible SAT instance {{iso_f}} takes
two sets of values and produces two outputs. The incoming values
labeled \textsf{inputs} are the inputs we need to test for
satisfiability. The other incoming values labeled \textsf{heap} are
the additional inputs needed to embed the original SAT instance {{f}}
into a reversible function. If these \textsf{heap} values are all
initialized to {{false}}, the output wire \textsf{satisfied?}
corresponds to the output that {{f}} would have produced on
\textsf{inputs}. The other outputs labeled \textsf{garbage} are not
needed for their own sake but they are important because they are used
as inputs to the adjoint of {{iso_f}} to reproduce the inputs exactly,
in anticipation of closing the loop with {{trace*}}.

To summarize, the top half of the circuit is the identity function
except that we have also managed to produce a boolean wire labeled
\textsf{satisfied?} that tells us if the inputs satisfy the desired
constraints. We can take this boolean value and use it to decide
whether to negate the bottom wire (labeled \textsf{control
  wire}). Specifically, if the inputs do \emph{not} satisfy {{f}}, the
control wire is negated. The last wire labeled \textsf{heap control
  wire} is negated if the heap values do not have the right initial
values, i.e., are not all {{false}}.

Let us call the above construction {{sat_f}}. If we now close the loop
using {{trace*}}, two things should happen:
\begin{itemize}
\item configurations in which the \textsf{heap} values are not all
  {{false}} will be annihilated;
\item configurations in which the \textsf{inputs} do not satisfy {{f}}
  will cause the \textsf{satisfied?} wire to be negated and hence will
  also be annihilated.
\end{itemize}
In other words, the only configurations that will survive are the ones in
which the \textsf{inputs} satisfy {{f}}. We simply need to arrange to
\emph{clone} these values and produce them as the output of the whole
circuit. The final construction is therefore:

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/sat3.pdf}
}
\end{center}

To make the previous discussion concrete, we present a small, but
complete, example. In our example, the SAT instance {{f}} is tiny: it
takes two inputs. This function is embedded into a reversible function
{{iso_f}} of type 
{{((bool * bool) * bool) <-> ((bool * bool) * bool)}} where the last
input represents the heap and the first two outputs represent the garbage. 
The realization of {{sat_f}} given below is parametrized by such 
a function {{iso_f}}. The inputs to {{sat_f}} are 
\textsf{heap control}, \textsf{control}, \textsf{heap}, \textsf{input-1}, and 
\textsf{input-2}. Its operation is simple: if the \textsf{heap} is {{true}}, 
\textsf{heap control} is negated, and if the last output of {{iso_f}} 
is {{false}}, \textsf{control} is negated:

%subcode{opsem} include main
%! columnStyle = rcl
% sat_f &:& ((((bool * bool) * bool) * bool) * bool) <-> ((((bool * bool) * bool) * bool) * bool)
% sat_f &=& ((swap* (*) id) (*) id) (;) 
% && ((assocl* (*) id) (*) id) (;) 
% &&   (((cnot (*) id) (*) id) (*) id) (;) 
% &&   assocr* (;) 
% &&   (assocr* (*) id) (;) 
% &&   (swap* (*) id) (;) 
% &&   assocr* (;) 
% &&   (id (*) assocl*) (;) 
% &&   (id (*) isof) (;) 
% &&   swap* (;) 
% &&   assocr* (;) 
% &&   (id (*) (id (*) swap*)) (;) 
% &&   (id (*) assocl*) (;) 
% &&   (id (*) ((inot (*) id) (*) id)) (;)
% &&   (id (*) (cnot (*) id)) (;) 
% &&   (id (*) ((inot (*) id) (*) id)) (;)
% &&   (id (*) assocr*) (;) 
% &&   (id (*) (id (*) swap*)) (;) 
% &&   assocl* (;)
% &&   swap* (;)
% &&   (id (*) Sym isof) (;) 
% &&   (id (*) assocr*) (;) 
% &&   assocl* (;) 
% &&   assocl*

Given the construction of {{sat_f}} we can build the full solver as
follows. The overall input is the cloning heap. The combinator given
to {{trace*}} takes the cloning heap and the inputs flowing around the
loop and produces two copies of these inputs. One copy is produced as
the overall output and another is fed back around the loop.

%subcode{opsem} include main
%! columnStyle = rcl
% solve_f &:& bool * bool <-> bool * Bool
% solve_f &=& trace* (
% && (assocr* (*) id) (;) 
% && assocr* (;)
% && (id (*) swap*) (;)
% && assocl* (;)
% && (assocr* (*) id) (;)
% && (clone2 (*) id) (;)
% && swap* (;)
% && (swap* (*) id) (;)
% && ((swap* (*) id) (*) id) (;)
% && assocl* (;)
% && assocl* (;)
% && (sat_f (*) id) (;)
% && (assocr* (*) id) (;)
% && (swap* (*) id) (;) 
% && ((id (*) swap*) (*) id) (;)
% && (assocl* (*) id) (;)
% && ((id (*) swap*) (*) id))

We can test our {{solve_f}} combinator using several SAT
instances. Here are two possible instances. The first instance is
satisfied by {{(false,false)}} and the
second is satisfied by {{(false,true)}} and {{(true,true)}}.

%subcode{opsem} include main
%! columnStyle = rcl
% iso_{f_1} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
% iso_{f_1} &=& (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
% && (((swap+ (*) id) (*) id) (;) toffoli (;) ((swap+ (*) id) (*) id)) (;) 
% && (id (*) swap+)
% 
% iso_{f_2} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
% iso_{f_2} &=& toffoli

It can indeed be verified using the semantics that the {{solve_f}}
combinators instantiated with the SAT instances produce the expected
results.

@@

>>>>

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that~$a$ and~$b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

@@

A major open problem remains, however: a higher-order extension of
{{Pi}}. This extension is of fundamental importance in all the originating
research areas. In physics, it allows a process or observer to be treated as
``data'' that can be transformed by higher-level processes or observed by
meta-level observers. In mathematics and logic, it allows the equivalence
between different proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed
as an isomorphism (of a higher type)
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code.

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{compact closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
as subtle (as we explain in the main body of the paper).  Our main novel
technical device to achieving the higher-order extension is a
\emph{fractional type}, which we so named because of its duality with
conventional product types.  By adding a formal (linear) dual to (type
theoretic) products, we can internalize the notion that ``taking an input''
corresponds to negative information, which turns out to be the basis of
first-class functions. The remainder of the paper reviews {{Pi}} and then
introduces the syntax and semantics of the extension with fractional
types. We then study the properties of the extended language and establish
its expressiveness via several constructions and examples that exploit its
ability to model higher-order computations.

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent Types include the unit type {{1}}, sum types {{b1 + b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary.  In particular we do not treat the above
values as first-class citizens.  They only occur when we want to observe the
effect of an isomorphism.

\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
(value) type isomorphisms.  They are also typed, by the shape of the (value)
type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
the following isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
\noindent Each line of the above table introduces a pair of dual
constants\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

Mathematically, {{Pi}} can be interpreted in the category of finite sets and
bijections: each value type denotes a finite set of a size calculated by
viewing the types as natural numbers and each combinator {{c : b1 <-> b2}}
denotes a bijection between the sets denoted by~{{b1}}
and~{{b2}}. Operationally, the semantics of {{Pi}} is given using two
mutually recursive interpreters: one going forward and one going
backwards. The use of {{sym}} switches control from one evaluator to the
other. We will present the operational semantics in the next section along
with the extension with fractional types and values. For now, we state
without proof that the evaluation of well-typed combinators always terminates
and that {{Pi}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

The relational semantics can be given an operational interpretation by
defining two functions: a forward evaluator and a backwards evaluator that
exercise the relation in opposite directions. It follows that if the forward
evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.

More generally, the semantics of {{PiTF}} using sets and relations, i.e., in
the category of finite sets and relations suggests a few additional immediate
properties.  The category of finite sets and relations is an example of a
\emph{biproduct dagger compact closed category} which are the categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value. We return to
such properties of {{PiTF}} in Sec.~\ref{sec:cat} after we present
programming examples that illustrate {{PiTF}} and its ability to express
various higher-order notions.

@@

In the following, we use {{bool}} as an abbreviation of {{1+1}} 
with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
abbreviation for {{right ()}}.

@@

As motivated, with fractional types, we can define a value that denotes a
first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
and a value {{v2 : b2}}. In other words, it corresponds to a function or
relation which when ``given'' a value {{v1 : b1}} ``releases'' the value 
{{v2 : b2}}. To emphasize this view, we introduce the abbreviation:

@@

As a simple example, consider the combinator {{swap+}} denoting boolean
negation.  It is straightforward to check that the combinator 
{{name swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:

%subcode{proof} include main
%@ ~
%@@ () ~~(name swap+)~~ (1/false,true)
%
%@ ~
%@@ () ~~(name swap+)~~ (1/true,false)

\noindent The relation {{name swap+}} maps {{()}} to a value which can be used
to convert any {{false}} to {{true}} and vice-versa. We formalize this
``apply'' combinator below.

@@

We are now a small step from implementing various higher-order
combinators that manipulate functions or relations. In particular, we
can \emph{apply} and \emph{compose} values representing relations as
shown below:

@@

%% \jacques{in the current paper, this seems to no longer make sense}
%% Yes I know, but I think we should leave it to explain the problem in Sec. 5
%% and motivate the different semantics

\begin{example}[Annihilation]
\label{ch3:ex;annihilation}
Consider the circuit below:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}
As motivated earlier, it is clearly impossible for the constraint implied by
{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
its negation. Thus there is no incoming value of type {{b}} that can ever go
through the circuit. Indeed, if we call the circuit above {{c}}, then we can
calculate that {{c}} does not relate {{v}} to anything. We call such a
circuit an annihilation circuit.
\end{example}

@@

Relational programming leverages set-theoretic relations and their
composition to express computational tasks in a declarative way. 
% For instance, consider this example:
% {{parent = {[ (A,B), (B,C), (A,D) ]} }}
% {{grandparent = parent (o) parent}}
% \noindent The example defines a relation {{parent}} specified using a
% set of tuples and another relation {{grandparent}} specified using the
% composition of two parent relations. If we wrote this example in a
% relational language (e.g., Prolog) and we executed the query
% {{grandparent(A)}}, we would get the answer {{ {[C]} }}.
It turns out that with the addition of the multiplicative trace, and
the move to relations motivated in the previous section, we can
express relational programming. We illustrate the idea using a
complete example.

@@

The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{trace* cInner}} is the desired combinator
{{c_R}}.  A little experimentation shows that we get the desired behavior if
as follows:
% %subcode{proof} include main
% %@ ~
% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
% %----
% %@ ~
% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
% %----
% %@ ~
% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
% %----
% %@ ~
% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
% %----
% %@ ~
% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
% %----
% %@ ~
% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
% %----
% %@ ~
% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
% %----
% %@ ~
% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
% In the first three lines, the first argument is a fixed point and hence the
% multiplicative trace would map the second input to the second output
% producing the desired relation. In the remaining five cases, the first
% argument is not a fixed point and hence all these cases would be rejected as
% solutions to the constraint imposed by the multiplicative trace. It simply
% remains to find the actual realization of {{cInner}} that would produce the
% behavior above.

@@

% The above example should convince the reader that the language {{Pi}}
% with multiplicative trace is expressive enough to model finite relational
% programming. 
% %%%%%%%%%%%%%%%
% \subsection{Solving Constraints}
% \label{ch3:sec:constraints}

% A large class of constraint satisfaction problems can be expressed
% using multiplicative traces. We illustrate the main ideas with the
% implementation of a SAT solver.

% In the usual setting, an instance of SAT is a function~{{f}} which,
% when given some boolean inputs, returns {{true}} or {{false}}. The
% function returns {{true}} when the inputs satisfy the constraints
% imposed by the structure of {{f}} and a solution to the SAT problem is
% the set of all inputs on which {{f}} produces {{true}}. The basic idea
% of our construction is to generalize the annihilation circuit from
% Ex.~\ref{ch3:ex;annihilation} to only annihilate values that fail to
% satisfy the constraints represented by the SAT instance {{f}}. To
% achieve this goal, we must however deal with several important
% details.

% First, because we are in a reversible world, our instance of SAT must be
% expressed as an isomorphism: this is easily achieved by the construction
% which embeds any boolean function {{f}} into a reversible one {{iso_f}} with
% a larger domain and range.  Given such a reversible function {{iso_f}} which
% represents a SAT instance, we first construct the circuit below:

% \begin{center}
% \scalebox{1.2}{
%   \includegraphics{diagrams/sat2.pdf}
% }
% \end{center}

% As shown in the circuit, the reversible SAT instance {{iso_f}} takes
% two sets of values and produces two outputs. The incoming values
% labeled \textsf{inputs} are the inputs we need to test for
% satisfiability. The other incoming values labeled \textsf{heap} are
% the additional inputs needed to embed the original SAT instance {{f}}
% into a reversible function. If these \textsf{heap} values are all
% initialized to {{false}}, the output wire \textsf{satisfied?}
% corresponds to the output that {{f}} would have produced on
% \textsf{inputs}. The other outputs labeled \textsf{garbage} are not
% needed for their own sake but they are important because they are used
% as inputs to the adjoint of {{iso_f}} to reproduce the inputs exactly,
% in anticipation of closing the loop with {{trace*}}.

% To summarize, the top half of the circuit is the identity function
% except that we have also managed to produce a boolean wire labeled
% \textsf{satisfied?} that tells us if the inputs satisfy the desired
% constraints. We can take this boolean value and use it to decide
% whether to negate the bottom wire (labeled \textsf{control
%   wire}). Specifically, if the inputs do \emph{not} satisfy {{f}}, the
% control wire is negated. The last wire labeled \textsf{heap control
%   wire} is negated if the heap values do not have the right initial
% values, i.e., are not all {{false}}.

% Let us call the above construction {{sat_f}}. If we now close the loop
% using {{trace*}}, two things should happen:
% \begin{itemize}
% \item configurations in which the \textsf{heap} values are not all
%   {{false}} will be annihilated;
% \item configurations in which the \textsf{inputs} do not satisfy {{f}}
%   will cause the \textsf{satisfied?} wire to be negated and hence will
%   also be annihilated.
% \end{itemize}
% In other words, the only configurations that will survive are the ones in
% which the \textsf{inputs} satisfy {{f}}. We simply need to arrange to
% \emph{clone} these values and produce them as the output of the whole
% circuit. The final construction is therefore:

% \begin{center}
% \scalebox{1.5}{
%   \includegraphics{diagrams/sat3.pdf}
% }
% \end{center}

% To make the previous discussion concrete, we present a small, but
% complete, example. In our example, the SAT instance {{f}} is tiny: it
% takes two inputs. This function is embedded into a reversible function
% {{iso_f}} of type 
% {{((bool * bool) * bool) <-> ((bool * bool) * bool)}} where the last
% input represents the heap and the first two outputs represent the garbage. 
% The realization of {{sat_f}} given below is parametrized by such 
% a function {{iso_f}}. The inputs to {{sat_f}} are 
% \textsf{heap control}, \textsf{control}, \textsf{heap}, \textsf{input-1}, and 
% \textsf{input-2}. Its operation is simple: if the \textsf{heap} is {{true}}, 
% \textsf{heap control} is negated, and if the last output of {{iso_f}} 
% is {{false}}, \textsf{control} is negated:

% %subcode{opsem} include main
% %! columnStyle = rcl
% % sat_f &:& ((((bool * bool) * bool) * bool) * bool) <-> ((((bool * bool) * bool) * bool) * bool)
% % sat_f &=& ((swap* (*) id) (*) id) (;) 
% % && ((assocl* (*) id) (*) id) (;) 
% % &&   (((cnot (*) id) (*) id) (*) id) (;) 
% % &&   assocr* (;) 
% % &&   (assocr* (*) id) (;) 
% % &&   (swap* (*) id) (;) 
% % &&   assocr* (;) 
% % &&   (id (*) assocl*) (;) 
% % &&   (id (*) isof) (;) 
% % &&   swap* (;) 
% % &&   assocr* (;) 
% % &&   (id (*) (id (*) swap*)) (;) 
% % &&   (id (*) assocl*) (;) 
% % &&   (id (*) ((inot (*) id) (*) id)) (;)
% % &&   (id (*) (cnot (*) id)) (;) 
% % &&   (id (*) ((inot (*) id) (*) id)) (;)
% % &&   (id (*) assocr*) (;) 
% % &&   (id (*) (id (*) swap*)) (;) 
% % &&   assocl* (;)
% % &&   swap* (;)
% % &&   (id (*) Sym isof) (;) 
% % &&   (id (*) assocr*) (;) 
% % &&   assocl* (;) 
% % &&   assocl*

% Given the construction of {{sat_f}} we can build the full solver as
% follows. The overall input is the cloning heap. The combinator given
% to {{trace*}} takes the cloning heap and the inputs flowing around the
% loop and produces two copies of these inputs. One copy is produced as
% the overall output and another is fed back around the loop.

% %subcode{opsem} include main
% %! columnStyle = rcl
% % solve_f &:& bool * bool <-> bool * Bool
% % solve_f &=& trace* (
% % && (assocr* (*) id) (;) 
% % && assocr* (;)
% % && (id (*) swap*) (;)
% % && assocl* (;)
% % && (assocr* (*) id) (;)
% % && (clone2 (*) id) (;)
% % && swap* (;)
% % && (swap* (*) id) (;)
% % && ((swap* (*) id) (*) id) (;)
% % && assocl* (;)
% % && assocl* (;)
% % && (sat_f (*) id) (;)
% % && (assocr* (*) id) (;)
% % && (swap* (*) id) (;) 
% % && ((id (*) swap*) (*) id) (;)
% % && (assocl* (*) id) (;)
% % && ((id (*) swap*) (*) id))

% We can test our {{solve_f}} combinator using several SAT
% instances. Here are two possible instances. The first instance is
% satisfied by {{(false,false)}} and the
% second is satisfied by {{(false,true)}} and {{(true,true)}}.

% %subcode{opsem} include main
% %! columnStyle = rcl
% % iso_{f_1} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
% % iso_{f_1} &=& (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
% % && (((swap+ (*) id) (*) id) (;) toffoli (;) ((swap+ (*) id) (*) id)) (;) 
% % && (id (*) swap+)
% % 
% % iso_{f_2} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
% % iso_{f_2} &=& toffoli

% It can indeed be verified using the semantics that the {{solve_f}}
% combinators instantiated with the SAT instances produce the expected
% results.

@@

\begin{comment}
Each type in {{PiTF}} is interpreted as a \emph{groupoid}. A groupoid can be
equivalently viewed as an oriented graph, a generalization of a group, or a
special category. We present the categorical view below.

Fractional types and values add considerable expressiveness to our
language. 
\end{comment}
